Variability in upwelling patterns across the Benguela

### 2019-10-13 Write_up/Paper1_writeup.Rmd
updated

Upwelling is a physical oceanographic process in which cold, nutrient rich waters are pushed to the surface as a result of the motion caused by alongshore equatorial winds. These upwelling events are known to have a large influence on the abundance, diversity, and distribution of marine organisms, and moreover, critically affect marine ecosystem processes (Add in references). Occurring globally throughout the world's oceans, four major eastern boundary upwelling systems (EBUS) can be recognised; each significantly impacting associated coastal systems. These respective systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun; @pauly]. There has been significant interest in major upwelling systems since the later 1960s, with major interest occuring off Peru [@walsh] and North-West Africa [@barber]. Research done by Copenhagen 1953, reported a number of sites along the west coast of southern Africa where upwelling was particularly intense. The two sites of intense upwelling activity were found off the Cape Peninsula and off Cape columbine [@copenhagen; @andrews]. Whilst upwelling plays a significant role in ecological processes, climate change largely influences this phenomenon.

Over the past three decades, anthropogenic climate change has established global concern as it has been shown to have negatively affected marine and terrestrial ecosystems [@hoegh]. The consequences of these changes are also shown to have had direct effects on the behaviour and functioning of marine organisms, thus ultimately affecting species trophic levels and integration [@doney]. According to the Bakun hypothesis, an increase in greenhouse gases will result in an increase in day-time warming and night-time cooling and ultimately cause an increase in temperature gradients which will then form stronger pressure gradients [@bakun]. Pressure gradients are known to affect wind patterns which will subsequently affect the intensity and duration of upwelling events [@hsieh; @mote; @lima]. Over the last few decades, climate change has resulted in drastic pressure and temperature gradients.

Several studies have provided significant evidence suggesting that there has been an increase in ocean surface temperatures over the last few decades (Add references; RWS: You can cite the IPCC for this). These temperatures have increased as a result of anthropogenically driven climate change [@gruber]. Warming is shown to slowly move down into the water column thus increasing thermal stratification (RWS: citation needed). This thermal stratification restricts the depth at which water is upwelled and therfore limits the amount of nutrients brought up to the surface with a decrease in oxygen and pH levels [@chhak; @jacox2011; @jacox2016]. Climate change impacting oceanic processes has major ecological consequences [@walther]. This is demonstrated in several regions increases in sea surface temperatures have been proven to greatly influence marine species survival, reproduction and abundance. Climate change has also been shown to largely influence coastal species such as corals, muscles and oysters [@hoegh]. Given the importance of primary production, it was determined that the Benguela upwelling system, the eastern boundary upwelling system in South Africa, should be the focus of this paper. Furthermore, it should be understood that increases in sea surface temperatures bear considerable economic impacts as well [@murawski2000; @murawski1993] therefore understanding the impact and cause of climate change has merit not only in mitigation efforts in limiting environmental degradation and maintaining ecosystem functioning, but in its negative affects to local and broader economies as well as human populations dependent on these coastal regions.

The 3000km South African coastline is bordered by the Agulhas current along the east coast and the Benguela current (an EBUS) along the west coast. These coastal regions display noticeable differences in seawater temperatures primarily due to the influences of the neighbouring ocean currents [@mead; @schlegel; @schlegel2016]. The Benguela Current is driven by the anticyclone high pressure systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region [@jarre; @hutchings]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@lutjeharms1987; @lutjeharms2003; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast [@luning1990; @lutjeharms2000; @hutchings; @schlegel].

The Benguela system is bounded by the equatorward and poleward ends by warm water regimes. These currents are known to be forced by local wind stress fields off southwest Africa [@nelson; @fennel]. This particular EBUS stretches along the south-western coast of Africa, from southern Angola to Cape Agulhas [@cole], with water temperatures coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@lutjeharms2000]. The upwelling period in the Benguela north of 25°C persists throughout the entire year but peaks during August. The Benguela upwelling region is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region in Namibia [@shannon; @cole]. Several studies have shown that extreme warming and cooling periods occur in the northern and southern Benguela systems [@stander; @walker; @shannon1984; @boyd; @brainard]. Furthermore, Walter (1937) found that warm periods of global ocean temperatures may have resulted as a lack of upwelling or poor upwelling events. During the periods of warming, the equatorward wind stress over the northern region of the Benguela system was stronger than usual yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region. In the southern Benguela region, continental shelf bathymetry and strong upwelling favourable winds are known to result in more intense upwelling events. Upwelling occurring along the southern coastline of southern Africa are usually generated by local winds.

The Cape peninsula is influenced by south-east trade winds during summer and westerlies during winter, thus a pronounced seasonal cycle is evident within patterns of wind movement across the coastline. During spring, summer, and autumn seasons, the anticyclonic high-pressure cell periodically ridges eastwards south of the continent (RWS: Citation). During this process, south easterly winds predominate, with peak frequencies occurring during spring and summer months resulting in more upwelling events during summer and winter months (RWS: Citation). During the winter seasons, low pressure cells are driven from the west to the east as a result of the atmospheric circumpolar waves and so north-westerly winds dominate during winter. South easterly winds dominate during summer but are still present for short durations during winter [@andrews]. Within a range of kilometres off the shore, wind conditions vary due to the effects of mountain funnelling within the region (RWS: Citation) and as a result conditions are variable in the nearshore and offshore.

@@

Strong temperature gradients existing between offshore and inshore oceanic waters may result in different upwelling event readings. Upwelling relatively close to the shore has been found to depend on the strength and direction of the wind and tend to vary based on seasonal wind variation [@andrews]. Inshore temperatures tend to be lower during summer months due to upwelling, but there is an increase in offshore temperatures in the later summer months as a result of warming of the South East Atlantic Surface waters [@andrews; @shannon]. Due to the temperature gradients that exist inshore and offshore, this study focused on observing upwelling patterns from the coastal region to 50km from the coastline.

>>>>

Upwelling is a physical oceanographic process in which cold, nutrient rich waters are pushed to the surface as a result of the motion caused by alongshore equatorial winds. These upwelling events are known to have a large influence on the abundance, diversity, and distribution of marine organisms, and moreover, critically affect marine ecosystem processes. Occurring globally throughout the world's oceans, four major eastern boundary upwelling systems (EBUS) can be recognised; each significantly impacting associated coastal systems. These respective systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun1990; @pauly; @bakun2015]. There has been significant interest in major upwelling systems since the later 1960s, with major interest occuring off Peru [@walsh] and North-West Africa [@barber]. Research done by Copenhagen 1953, reported a number of sites along the west coast of southern Africa where upwelling was particularly intense. The two sites of intense upwelling activity were found off the Cape Peninsula and off Cape columbine [@copenhagen; @andrews]. Whilst upwelling plays a significant role in ecological processes, climate change largely influences this phenomenon.

Over the past three decades, anthropogenic climate change has established global concern as it has been shown to have negatively affected marine and terrestrial ecosystems [@hoegh]. The consequences of these changes are also shown to have had direct effects on the behaviour and functioning of marine organisms, thus ultimately affecting species trophic levels and integration [@doney]. According to the Bakun hypothesis, an increase in greenhouse gases will result in an increase in day-time warming and night-time cooling and ultimately cause an increase in temperature gradients which will then form stronger pressure gradients [@bakun1990]. Pressure gradients are known to affect wind patterns which will subsequently affect the intensity and duration of upwelling events [@hsieh; @mote; @lima]. Over the last few decades, climate change has resulted in drastic pressure and temperature gradients.

Several studies have provided significant evidence suggesting that there has been an increase in ocean surface temperatures over the last few decades [@vose2012noaa]. These temperatures have increased as a result of anthropogenically driven climate change [@gruber]. Warming is shown to slowly move down into the water column thus increasing thermal stratification [@chhak; @jacox2011]. This thermal stratification restricts the depth at which water is upwelled and therfore limits the amount of nutrients brought up to the surface with a decrease in oxygen and pH levels [@chhak; @jacox2011; @jacox2016]. Climate change impacting oceanic processes has major ecological consequences [@walther]. This is demonstrated in several regions increases in sea surface temperatures have been proven to greatly influence marine species survival, reproduction and abundance. Climate change has also been shown to largely influence coastal species such as corals, muscles and oysters [@hoegh]. Given the importance of primary production, it was determined that the Benguela upwelling system, the eastern boundary upwelling system in South Africa, should be the focus of this paper. Furthermore, it should be understood that increases in sea surface temperatures bear considerable economic impacts as well [@murawski1993; @murawski2000] therefore understanding the impact and cause of climate change has merit not only in mitigation efforts in limiting environmental degradation and maintaining ecosystem functioning, but in its negative affects to local and broader economies as well as human populations dependent on these coastal regions.

The 3000km South African coastline is bordered by the Agulhas current along the east coast and the Benguela current (an EBUS) along the west coast. These coastal regions display noticeable differences in seawater temperatures primarily due to the influences of the neighbouring ocean currents [@mead; @schlegel2016; @schlegel]. The Benguela Current is driven by the anticyclone high pressure systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region [@jarre; @hutchings]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@lutjeharms1987; @lutjeharms2003; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast [@luning1990; @lutjeharms2000; @hutchings; @schlegel].

The Benguela system is bounded by the equatorward and poleward ends by warm water regimes. These currents are known to be forced by local wind stress fields off southwest Africa [@nelson; @fennel]. This particular EBUS stretches along the south-western coast of Africa, from southern Angola to Cape Agulhas [@cole], with water temperatures coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@lutjeharms2000]. The upwelling period in the Benguela north of 25°C persists throughout the entire year but peaks during August. The Benguela upwelling region is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region in Namibia [@shannon1984; @shannon1986; @cole]. Several studies have shown that extreme warming and cooling periods occur in the northern and southern Benguela systems [@stander; @walker; @shannon1984; @boyd; @brainard]. Furthermore, Walter (1937) found that warm periods of global ocean temperatures may have resulted as a lack of upwelling or poor upwelling events. During the periods of warming, the equatorward wind stress over the northern region of the Benguela system was stronger than usual yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region. In the southern Benguela region, continental shelf bathymetry and strong upwelling favourable winds are known to result in more intense upwelling events. Upwelling occurring along the southern coastline of southern Africa are usually generated by local winds.

The Cape peninsula is influenced by south-east trade winds during summer and westerlies during winter, thus a pronounced seasonal cycle is evident within patterns of wind movement across the coastline. During spring, summer, and autumn seasons, the anticyclonic high-pressure cell periodically ridges eastwards south of the continent [@preston1977towards; @bakun1990]. During this process, south easterly winds predominate, with peak frequencies occurring during spring and summer months resulting in more upwelling events during summer and winter months [@bakun1990; bakun2015]. During the winter seasons, low pressure cells are driven from the west to the east as a result of the atmospheric circumpolar waves and so north-westerly winds dominate during winter. South easterly winds dominate during summer but are still present for short durations during winter [@andrews]. Within a range of kilometres off the shore, wind conditions vary due to the effects of mountain funnelling within the region [@andrews] and as a result conditions are variable in the nearshore and offshore.

@@

Strong temperature gradients existing between offshore and inshore oceanic waters may result in different upwelling event readings. Upwelling relatively close to the shore has been found to depend on the strength and direction of the wind and tend to vary based on seasonal wind variation [@andrews]. Inshore temperatures tend to be lower during summer months due to upwelling, but there is an increase in offshore temperatures in the later summer months as a result of warming of the South East Atlantic Surface waters [@andrews; @shannon1984]. Due to the temperature gradients that exist inshore and offshore, this study focused on observing upwelling patterns from the coastal region to 50km from the coastline.
